---
title: "Cars"
author: "Marko Intihar"
date: "10/03/2020"
output: 
  html_document:
    css: template1.css
    number_sections: yes
    toc: yes
    toc_float: yes
---

# Introduction

This report covers inspection of the **cars dataset** (**mpg**) from the **ggplot2** library. The content covers:

* quick dataset exploration
* exploratory data analysis
* car highway miles consumption prediction modelling


# The Dataset

```{r data, echo=FALSE, include=FALSE}
library(dplyr)
library(tidyr)
library(ggplot2)
library(cowplot)
library(hexbin)

library(kableExtra)

# data
df <- ggplot2::mpg
str(df) # table structure

df <- df %>% 
  mutate(manufacturer = as.factor(manufacturer),
         model = as.factor(model),
         trans = as.factor(trans),
         drv = as.factor(drv),
         fl = as.factor(fl),
         class = as.factor(class),
         cyl = as.factor(cyl))

```

Dataset preview:

```{r data look, echo=TRUE, include=TRUE}
# structure
str(df)

# summary
summary(df)

# data frame dimensions
nrow(df)
ncol(df)

# pre-look
head(df)

```


# Exploratory Data Analysis (EDA)

In this report we try to explore the given cars dataset. By considering:

* each individual variable;
* two variables and their connection;
* many variables;


## Explore each individual variable

Dataset includes given variables:

* **manufacturer** - car manufacturer;
* **model** - car model;
* **displ** - engine displacement in litres;
* **year** - year of manufacture;
* **cyl** - number of cylinders;
* **trans** - type of transmission;
* **drv** - type of wheel drive;
* **cty** - city miles per gallon consumption;
* **hwy** - highway miles per gallon consumption;
* **fl** - fuel type;
* **class** - type of car;

Figure below shows distribution of variables **hwy** and **cty**.

```{r ctyhwy, fig.align='center', fig.width=12, fig.height=10, fig.show='asis', echo=FALSE}
# cty histogram for exporting
fig.cty <- df %>% 
  ggplot(aes(x = cty)) +
  geom_histogram(binwidth = 1, color = "black", fill = "gray40") +
  scale_x_continuous(breaks = seq(0,50,2)) +
  scale_y_continuous(breaks = seq(0,30,2)) +
  xlab("City miles per gallon") +
  ylab("Frequency") +
  ggtitle("City and Highway miles per gallon - histograms") +
  theme(axis.text = element_text(size = 14),
        axis.title = element_text(size = 16),
        plot.title = element_text(size = 20, face = "bold"))

# hwy histogram for exporting
fig.hwy <- df %>% 
  ggplot(aes(x = hwy)) +
  geom_histogram(binwidth = 1, color = "black", fill = "gray40") +
  scale_x_continuous(breaks = seq(0,50,2)) +
  scale_y_continuous(breaks = seq(0,40,2)) +
  xlab("Highway miles per gallon") +
  ylab("Frequency") +
  theme(axis.text = element_text(size = 14),
        axis.title = element_text(size = 16),
        plot.title = element_text(size = 20, face = "bold"))

cowplot::plot_grid(fig.cty, fig.hwy, ncol = 1)

```

Distribution of continuous variable **displ** is presented with a frequency polygon.


```{r displ, fig.align='center', fig.width=12, fig.height=8, fig.show='asis', echo=FALSE}
df %>% 
  ggplot(aes(x = displ)) +
  geom_freqpoly(binwidth = 0.2, size = 1.2, color = "black") +
  scale_x_continuous(breaks = seq(0,8,0.2)) +
  scale_y_continuous(breaks = seq(0,40,5)) +
  xlab("Displacement in litres") +
  ylab("Frequency") +
  ggtitle("Engine displacement - frequency polygon") +
  theme(axis.text = element_text(size = 14),
        axis.title = element_text(size = 16),
        plot.title = element_text(size = 20, face = "bold"))

```

Selected discrete variables are drawn using bar plot.

```{r discrete, fig.align='center', fig.width=12, fig.height=12, fig.show='asis', echo=FALSE}

fig.manufact <- df %>% 
  ggplot(aes(x = manufacturer)) +
  geom_bar(stat = "count", fill = "gray60", color = "black") +
  scale_y_continuous(breaks = seq(0,50,5)) +
  xlab("Car manufacturer") +
  ylab("Car count") +
  theme(axis.text = element_text(size = 14),
        axis.title = element_text(size = 16),
        plot.title = element_text(size = 20, face = "bold"))

fig.cyl <- df %>% 
  ggplot(aes(x = cyl)) +
  geom_bar(stat = "count", fill = "gray60", color = "black") +
  scale_y_continuous(breaks = seq(0,90,10)) +
  xlab("Number of cylinders") +
  ylab("Car count") +
  theme(axis.text = element_text(size = 14),
        axis.title = element_text(size = 16),
        plot.title = element_text(size = 20, face = "bold"))

fig.trans <- df %>% 
  ggplot(aes(x = trans)) +
  geom_bar(stat = "count", fill = "gray60", color = "black") +
  scale_y_continuous(breaks = seq(0,90,10)) +
  xlab("Transmission type") +
  ylab("Car count") +
  theme(axis.text = element_text(size = 14),
        axis.title = element_text(size = 16),
        plot.title = element_text(size = 20, face = "bold"))

fig.fl <- df %>% 
  ggplot(aes(x = fl)) +
  geom_bar(stat = "count", fill = "gray60", color = "black") +
  scale_y_continuous(breaks = seq(0,200,25)) +
  xlab("Fuel type") +
  ylab("Car count") +
  theme(axis.text = element_text(size = 14),
        axis.title = element_text(size = 16),
        plot.title = element_text(size = 20, face = "bold"))

first_row <- cowplot::plot_grid(fig.manufact, nrow = 1)
second_row <- cowplot::plot_grid(fig.trans, nrow = 1)
third_row <- cowplot::plot_grid(fig.fl, fig.cyl, nrow = 1)

cowplot::plot_grid(first_row, second_row, third_row, 
                                  ncol = 1, nrow = 3)

```


## Explore two variables


Now we will compare **hwy** with selected variables. Let's begin by comparing **hwy** with **cty**.

```{r hwycty, fig.align='center', fig.width=12, fig.height=8, fig.show='asis', echo=FALSE}
df %>% 
  ggplot(aes(x = cty, y = hwy)) +
  geom_point(size = 5, color = "black") +
  geom_smooth(method = "lm", se = T) +
  scale_x_continuous(breaks = seq(0,40,2.5)) +
  scale_y_continuous(breaks = seq(0,50,2.5)) +
  xlab("City miles per gallon") +
  ylab("Highway miles per gallon") +
  ggtitle("City VS Highway miles per gallon - scatter plot") +
  theme(axis.text = element_text(size = 16),
        axis.title = element_text(size = 16),
        plot.title = element_text(size = 20, face = "bold"),
        panel.border = element_rect(color = "black", fill = NA, size = 1.2))

```

If we compare **hwy** with **displ** using hexagonal heatmap.

```{r hwydispl, fig.align='center', fig.width=12, fig.height=8, fig.show='asis', echo=FALSE}
df %>% 
  ggplot(aes(y = hwy, x = displ)) +
  geom_hex(binwidth = c(0.25,2)) +
  scale_x_continuous(breaks = seq(0,10,0.5)) +
  scale_y_continuous(breaks = seq(0,50,5)) +
  scale_fill_viridis_c(option = "magma") +
  xlab("Engine displacement (in litres)") +
  ylab("Highway miles per gallon") +
  ggtitle("Highway miles VS engine displacement - Hexagonal heatmap 2D") +
  theme(axis.text = element_text(size = 14),
        axis.title = element_text(size = 16),
        plot.title = element_text(size = 18, face = "bold"),
        legend.text = element_text(size = 14),
        legend.title = element_text(size = 16),
        panel.border = element_rect(color = "black", fill = NA, size = 1.2))

```

**hwy** with **manufacturer**:

```{r hwymanufact, fig.align='center', fig.width=12, fig.height=8, fig.show='asis', echo=FALSE}
df %>% 
  ggplot(aes(y = hwy, x = manufacturer, fill = manufacturer)) +
  geom_boxplot(size = 1.3, 
               outlier.size = 5) +
  geom_jitter(color = "gray40", size = 2) +
  scale_y_continuous(breaks = seq(0,50,5)) +
  scale_fill_viridis_d(option = "plasma") +
  xlab("Car manufacturer") +
  ylab("Highway miles per gallon") +
  ggtitle("Highway miles VS car manufacturer - boxplot") +
  coord_flip() +
  theme(axis.text = element_text(size = 14),
        axis.title = element_text(size = 16),
        plot.title = element_text(size = 18, face = "bold"),
        panel.border = element_rect(color = "black", fill = NA, size = 1.2),
        legend.position = "none")

```

**hwy** with **class**:

```{r hwyclass, fig.align='center', fig.width=12, fig.height=8, fig.show='asis', echo=FALSE}
df %>% 
  ggplot(aes(y = hwy, x = class, fill = class)) +
  geom_violin() +
  geom_jitter(color = "gray40", size = 2) +
  scale_y_continuous(breaks = seq(0,50,5)) +
  scale_fill_viridis_d(option = "viridis") +
  xlab("Car type") +
  ylab("Highway miles per gallon") +
  ggtitle("Highway miles VS car type - violin plot") +
  theme(axis.text = element_text(size = 14),
        axis.title = element_text(size = 16),
        plot.title = element_text(size = 18, face = "bold"),
        panel.border = element_rect(color = "black", fill = NA, size = 1.2),
        legend.position = "none")

```


## Explore many variables

Figure below uses different variables for many plot elements:

* **cty** mapped on *x axis*
* **hwy** mapped on *y axis*
* **cyl** mapped on dot color
* **drv** mapped on shape of dots
* **fl** used for creating facets

```{r manyvars, fig.align='center', fig.width=12, fig.height=8, fig.show='asis', echo=FALSE}
df %>% 
  ggplot(aes(x = cty, y = hwy, color = cyl, shape = drv)) +
  geom_point(position = "jitter", size = 5) +
  facet_wrap(vars(fl), labeller = "label_both") +
  scale_size(range = c(0.01,10)) +
  scale_x_continuous(breaks = seq(0,60,5)) +
  scale_y_continuous(breaks = seq(0,60,5)) +
  xlab("City miles per gallon") +
  ylab("Highway miles per gallon") +
  ggtitle("Hwy ~ Cty ~ cyl ~ drv ~ fl - scatter plot") +
  theme_bw() +
  theme(axis.text = element_text(size = 16),
        axis.title = element_text(size = 16),
        plot.title = element_text(size = 20, face = "bold"),
        legend.title = element_text(size = 16),
        legend.text =  element_text(size = 12),
        panel.border = element_rect(color = "black", fill = NA, size = 1.2),
        strip.background = element_rect(colour = "black", fill = "white"),
        strip.text = element_text(size = 14))

```


# Highway miles prediction models

In the final section of the report we would like to build models for car highway miles consumption predictions. Each model will use different variables as predictors (different number of variables and different types of variables). Some guildelines in model building:

* target variable (dependant variable): **hwy** 
* selected predictor variables (independantd variables): **cty**, **displ**, **cyl**, **drv**, **fl**
* prediction model: **linear regression model**
* cross-validation applied: 80% of dataset used for **training**, 20% dataset used for **testing**
* model performance measure: **Root Means Squared Error** (**RMSE**)

**RMSE** is calculated as:

$$RMSE = \sqrt{\frac{1}{n}\sum_{i=1}^{n}(\hat{y}_i-y_i)^2}$$

where $y_i$ is actual highway consumption of car $i$ and $\hat{y}_i$ is predicted highway consumption value of car $i$.


```{r modelbuilding, echo=FALSE, include=FALSE}

# Functions needed
source("C:/Users/Marko/Desktop/project/functions.R")

# Generate df for models
predictor.vars <- c("cty", "displ", "cyl", "drv", "fl")
predicted.var <- "hwy"
cars.models <- generate.models(predictors.vars = predictor.vars, outcome.var = predicted.var)

# Split df
set.seed(123)
df <- split.sample(df = df, train.size = 0.8)
df.train <- df %>% filter(sample == "train")
df.test <- df %>% filter(sample == "test")

# Fit models and calculate RMSE

cars.models <- cars.models %>% # add RMSE column
  mutate(RMSE = NA)

for(id.model in 1:nrow(cars.models)){ # loop over each model
  
  # train model
  formula <- cars.models[id.model,"formula"]
  lm.model <- lm(formula = formula, data = df.train)
  
  # predict price (test dataset)
  df.test <- df.test %>% 
    mutate(`hwy predicted` = predict(lm.model, .)) 
  
  # calculate RMSE (predicted price VS actual price)
  SSE <- (df.test$`hwy predicted` - df.test$hwy)^2 # sum of squared errors
  RMSE <- sqrt(mean(SSE))
  
  # write RMSE back to table
  cars.models[id.model,"RMSE"] <- RMSE
}


```

In the table below we show every model we have built, number of model's predictors, predictors and model performance.

```{r modelstable, include=TRUE, echo=FALSE}
cars.models %>% 
  select(`nr. predictors`, predictors, RMSE) %>% 
  kable() %>%
  kable_styling(full_width = F)
```

The figure below shows performance of each model.

```{r modelperformance, fig.align='center', fig.width=12, fig.height=8, fig.show='asis', echo=FALSE}

predictors.levels <- cars.models %>% arrange(id) %>% pull(predictors) # levels for predictors factor variable

cars.models %>% 
  mutate(`nr. predictors` = as.factor(as.character(`nr. predictors`)),
         predictors = factor(predictors, levels = predictors.levels)) %>% 
  ggplot(aes(x = predictors, y = RMSE, fill = `nr. predictors`)) +
  geom_bar(stat = "identity", color = "black") +
  scale_y_continuous(breaks = seq(0, 5, 0.25)) +
  scale_fill_viridis_d(option = "inferno", direction = -1) +
  xlab("Predictors") +
  ylab("RMSE") +
  ggtitle("Highway miles prediction model performance") +
  theme(axis.text = element_text(size = 12),
        axis.text.x = element_text(angle = 90, hjust = 1),
        axis.title = element_text(size = 16),
        plot.title = element_text(size = 16, face = "bold"),
        legend.title = element_text(size = 14),
        legend.text = element_text(size = 12),
        panel.border = element_rect(color = "black", fill = NA, size = 1.5))

```




